Bumble bees and wildflowers along a mountain slope exhibit nonlinear patterns of abundance and turnover, punctuated by the tree line ecotone

### 2021-02-18 ms_1/ms_1.Rmd
ms

Bumble bees (Hymenoptera: *Bombus* spp.) are both a classic model organism in community ecology and an emergent conservation priority due to recent declines [@Goulson2008-oe] and manifest vulnerability to climate change [@Soroye2020-zb]. The genus *Bombus* is believed to have arisen in the mountains of Asia during the cooling climate of the Eocene-Oligocene boundary (~34 mya), and its subsequent spread and diversification along mountain corridors and into lowland habitats was likely driven by alternating range expansion during cooling periods and retreat to higher elevations during warming periods [@Hines2008-er; @Stewart2010-jn; @Martinet2018-rb]. Today, bumble bees remain characteristically cold-adapted species [@Heinrich1994-cv], with peak abundance and diversity associated with mountain ranges and northern latitudes, where they are often the dominant guild of flower-visiting insects and the principal pollinators of entomophilous flora [@Williams1998-xi; @Goulson2010-lo]. Climate-induced range shifts are expected to be especially pronounced in the mountain habitats that host the core of bumble bee biodiversity, together with their floral mutualists, and upward movement of bumble bee species has already been documented [@Ploquin2013-da; @Kerr2015-dg; @Pyke2016-ll; @Fourcade2019-ct; @Soroye2020-zb; @Marshall2020-pc]. Populations of bumble bee species already restricted to the highest elevation zones can be expected to suffer geometric habitat shrinkage and increasing genetic isolation as they move toward mountain peaks; the upward advance of lowland species may induce novel competitive pressures and disrupt the pollination networks in which bumble bees participate [@Brosi2013-mu; @Ishii2013-sm]; and concurrent elevational [@Lenoir2008-yn] and/or phenological [@Pyke2016-ll] shifts in the floral community could generate mismatches between bumble bee species and their historic floral hosts.

The outcomes of range shifts in mountain bumble bee and floral communities have obvious significance from a conservation perspective, but they also provide a model system in which to study the general phenomenon of community assembly under climate change, a phenomenon of both historical interest and prospective urgency. The goal of anticipating, interpreting, and potentially mitigating the effects of ranges shifts, however, is predicated upon understanding the current elevational and temporal patterns of bumble bee communities and those of their floral hosts. In particular, it is important to detect nonlinearities in species-, community-, and interaction-level responses to elevational climate gradients, such as the floristic threshold effect that has been documented at the tree line ecotone [@Descombes2017-id].

In the present study, we approach this topic at three levels of biological organization. First, we ask how mountain bumble bee species are distributed in elevation and intra-annual time with respect to one another and with respect to their floral hosts. Then, we move from the level of individual species to that of community composition and explore the structure of species turnover (i.e. β-diversity) of bumble bees and flora in response to elevation. Finally, propagate the question to the level of species interactions, investigating the turnover of bee-flower interaction partners through elevation. In each of these analyses, we employ flexible modeling techniques capable of capturing nonlinear responses to elevation and time with the goal of identifying where, when, and for which species the impacts of climate-induced range shifts should be expected to be most acute.

@@

The floristic importance of the tree line ecotone has been noted in previous work [@Pellissier2010-kg; @Descombes2017-id]. By studying the floral community through the lens of foraging bumble bees, we provide a functional extension of this floristic pattern, showing that the tree line ecotone both a boundary above which floral resource availability sharply declines and a critical distributional interface between high-elevation bumble bees and their habitat-generalist counterparts. Similar patterns with respect to the tree line ecotone have been described for bumble bees and their floral hosts in the Mediterranean system of Mount Olympus [@Minachilis2020-xd], suggesting that these findings are a general property of mountain ecosystems.

@@

A salient question that could not be answered by our sampling approach is whether mountain bumble bees adaptively forage up- or downslope, as suggested by @Lundberg1980-qk. Bumble bees have large foraging ranges and have been shown both to track resources through space and time [@Devoto2014-ay] and to cross forest matrix to reach patches of foraging habitat [@Mola2020-yc]. Assuming an average slope of 20° and a foraging range of 1 km, a bumble bee could travel up- or downslope by more than 340 m, thus spanning an elevation belt nearly 700 m wide. Such 3-dimensional foraging would enable bumble bees both to exploit the elevational turnover of floral species and to track preferred species through their elevationally staggered phenology, the latter constituting a sort of physiological time travel [@Van_Straalen1983-df.

he rate of β-diversity --- whether that of bumble bees, flora, or interactions --- accelerates sharply near the tree line ecotone (~1500 m) and continues accelerating for the remainder of the elevation gradient (Figure 8). Thus, a linear elevation gradient generates a nonlinear biological response, punctuated by the tree line ecotone, above which further increase in elevation becomes an increasingly stringent filter of community assembly. Such patterns have been described previously for plants (Pellissier et al. 2010, Descombes et al. 2017); our study suggests they extend also to bumble bees, and that they are propagated from the level of community composition to that of species interactions between bumble bees and their floral hosts.

The central finding of our study, evident in each of our analyses, is that a linear elevation gradient can generate nonlinear biological responses in terms of both abundance and β-diversity, and the latter at both at the level of community composition and at the level of species interactions. While elevational climate variation cannot be completely disentangled from other variables that covary with elevation [@Hodkinson2005-af], nonlinear responses of plants and pollinators along an elevational climate gradient suggest that we might expect similar nonlinearities in response to temporal climate change. Indeed, such nonlinear responses to climate have already been reported, including a recent study that documented nonlinear decline of insect abundance over a 37-year temporal climate gradient [@Lister2018-jp].

Independent of their utility as space-for-time climate proxies, mountains per se are unique ecological theaters in which extrinsic temporal climate change interacts with an intrinsic elevational climate gradation to alter the spatial and temporal distributions of species and interactions [@Telwala2013-sc; @CaraDonna2014-nq; @Miller-Struttmann2014-wz; @Rafferty2020-aj]. The nonlinearity of abundance observed in our study warns that temporal climate change on mountain slopes should be expected to involve the redistribution of discrete peaks of species abundance.

It should also be expected that climate change on mountain slopes will be punctuated by the tree line ecotone, which constitutes, at least with respect to bumble bees and wildflowers, a region of unique resource abundance and a transition to high and accelerating turnover of species and interactions [@@Descombes2017-id]. The observation that the upslope advancement of alpine tree lines often lags behind climate warming [@Dullinger2004-qu] suggests a danger that high elevation florivores could move upslope faster than their floral resources, or perhaps be forced to cope with increased competitive pressure from other florivore species formerly restricted to lower elevations [@Hoiss2012-sn].

>>>>

Bumble bees (Hymenoptera: *Bombus* spp.) are both a classic model organism in community ecology and an emergent conservation priority due to recent declines [@Goulson2008-oe] and manifest vulnerability to climate change [@Soroye2020-zb]. The genus *Bombus* is believed to have arisen in the mountains of Asia during the cooling climate of the Eocene-Oligocene boundary (~34 mya), and its subsequent spread and diversification along mountain corridors and into lowland habitats was likely driven by alternating range expansion during cooling periods and retreat to higher elevations during warming periods [@Hines2008-er; @Stewart2010-jn; @Martinet2018-rb]. Today, bumble bees remain characteristically cold-adapted species [@Heinrich1994-cv], with peak abundance and diversity associated with mountain ranges and northern latitudes, where they are often the dominant guild of flower-visiting insects and the principal pollinators of entomophilous flora [@Williams1998-xi; @Goulson2010-lo]. Climate-induced range shifts are expected to be especially pronounced in the mountain habitats that host the core of bumble bee biodiversity, together with their floral mutualists, and upward movement of bumble bee species has already been documented [@Ploquin2013-da; @Kerr2015-dg; @Pyke2016-ll; @Fourcade2019-ct; @Soroye2020-zb; @Marshall2020-pc]. Populations of bumble bee species already restricted to the highest elevation zones can be expected to suffer geometric habitat shrinkage and increasing genetic isolation as they move toward mountain peaks; the upward advance of lowland species may introduce novel competitive pressures and disrupt the pollination networks in which bumble bees participate [@Brosi2013-mu; @Ishii2013-sm]; and concurrent elevational [@Lenoir2008-yn] and/or phenological [@Pyke2016-ll] shifts in the floral community could generate mismatches between bumble bee species and their historic floral hosts.

The outcomes of range shifts in mountain bumble bee and floral communities have obvious significance from a conservation perspective, but they also provide a model system in which to study the general phenomenon of community assembly under climate change, a phenomenon of both historical interest and prospective urgency. Anticipating, interpreting, and potentially mitigating the effects of ranges shifts, however, necessitate an understanding the current elevational and temporal patterns of bumble bee communities and those of their floral hosts that will be acted upon by climate change. In particular, it is important to detect nonlinearities in species-, community-, and interaction-level responses to elevational climate gradients, such as the floristic threshold effect that has been documented at the tree line ecotone [@Descombes2017-id].

In the present study, we approach this topic at three levels of biological organization. First, we ask how mountain bumble bee species are distributed in elevation and intra-annual time with respect to one another and with respect to their floral hosts. Then, we move from the level of individual species to that of community composition and explore the structure of species turnover (i.e. β-diversity) of bumble bees and flora in response to elevation. Finally, we move to the level of species interactions, investigating the turnover of bee-flower interaction partners through elevation. In each of these analyses, we employ flexible modeling techniques capable of capturing nonlinear responses with the goal of identifying where, when, and for which species the impacts of climate-induced range shifts should be expected to be most acute.

@@

- Recap the point of the study

The core finding of our study, evident in each of our analyses, is that a linear elevation gradient can generate nonlinear biological responses in terms of both abundance and β-diversity, and the latter both at the level of community composition and at the level of species interactions.

The most salient pattern in our study is the unique importance of the elevation band falling at about 1500 m, which corresponds to the lower edge of the tree line ecotone in the Berchtesgadener Alps [@Kostler1970-ya; @Mayer1970-zn] and specifically that of the elevational transects used in our study. The floristic importance of the tree line ecotone has been noted in previous work [@Pellissier2010-kg; @Descombes2017-id]. By studying the floral community through the lens of foraging bumble bees, however, we provide a functional extension of this floristic pattern, demonstrating that the tree line ecotone comprises (1) a boundary above which floral resource availability sharply declines, (2) a distributional interface between high-elevation bumble bees and their low-elevation or habitat-generalist counterparts, and (3) a zone of rapid acceleration of species and interaction β-diversity. The significance of the tree line ecotone as a threshold of floral resource availability and as a zone of overlapping bumble bee distributions corroborates the recent findings of @Minachilis2020-xd in the Mediterranean system of Mount Olympus, indicating that these patterns are likely to generalize well across mountain ecosystems. Our finding that interaction β-diversity is also punctuated by the tree line ecotone follows naturally from these compositional patterns, and we expect that the β-diversity patterns found in our system will prove similarly generalizable.

The observation that the upslope advancement of alpine tree lines often lags behind climate warming [@Dullinger2004-qu] suggests a danger that high elevation bumble bees could move upslope faster than their floral resources. Indeed, a somewhat unexpected finding of our study is that high-elevation bumble bee species (*B. mucidus*, *B. monticola*, *B. mendax*, and *B. pyrenaeus*) were associated with floral hosts that peaked in abundance at much lower elevations, largely outside the range of their bumble bee visitors. As the upper boundary of high floral resource abundance, the lower edge of the tree line ecotone likely functions as critical foraging habitat for high-elevation bumble bees. Future warming could force these species farther into resource-poor alpine grasslands while at the same time increasing resource competition at the tree line as dominant low-elevation species, like *B. pascuorum*, extend their ranges upslope.

Our analyses of species and interaction β-diversity reveal a striking contrast between a relative stable bumble bee community and an extremely dynamic floral community. The dominance of total interaction β-diversity by floral species turnover is consistent with the findings of @Simanonok2014-ou in the Rocky Mountains of North America, and it is perhaps a predictable relationship given the exceptional cold-hardiness of bumble bees [@Heinrich1994-cv] and the sensitivity of plants to temperature gradients []. Nevertheless, it highlights the behavioral challenge that bumble bees have presumably faced for the entirety of their evolutionary history [] and that may explain the remarkable flexibility and intelligence of their flower handling []: how to forage efficiently in a diverse that turns over rapidly in both space and time. A fascinating question that could not be answered by our sampling approach is whether mountain bumble bees adaptively forage up- or downslope, as suggested by @Lundberg1980-qk. Bumble bees have large foraging ranges and have been shown both to track resources through space and time [@Devoto2014-ay] and to cross forest matrix to reach patches of foraging habitat [@Mola2020-yc]. Assuming an average slope of 20° and a foraging range of 1 km, a bumble bee could travel up- or downslope by more than 340 m, thus spanning an elevation belt nearly 700 m wide. Such 3-dimensional foraging would enable bumble bees both to exploit the elevational turnover of floral species and to track preferred species through their elevationally staggered phenology, the latter constituting a sort of physiological time travel [@Van_Straalen1983-df.

- What do these findings tell use about climate change and range shifts in mountain ecosystems more generally?
- Do our results suggest any management strategies?

As in the past, so now are mountains the primarily theaters of ecological change for bumble bees.

Mountains unique ecological theaters in which extrinsic temporal climate change interacts with an intrinsic elevational climate gradation to alter the spatial and temporal distributions of species and interactions [@Telwala2013-sc; @CaraDonna2014-nq; @Miller-Struttmann2014-wz; @Rafferty2020-aj].

It should also be expected that climate change on mountain slopes will be punctuated by the tree line ecotone, which constitutes, at least with respect to bumble bees and wildflowers, a region of unique resource abundance and a transition to high and accelerating turnover of species and interactions [@@Descombes2017-id].

Retreating to higher elevation in response to climatic warming is not a new phenomenon in the evolutionary history of bumble bees

@@
